### 2013-07-02 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
adding titan info and communication parameter explanation

\subsection{Titan }
\label{subsec:titan}

@@

\begin{figure}[htpb!]
  \begin{center}
    \includegraphics[width=6in]{chapters/parallel_mc/titan_comm_parameters.pdf}
  \end{center}
  \caption{\textbf{Neumann Ulam parallel algorithm free parameter
      sensitivity.}  \textit{All runtimes were within 6.5\% of the
      fastest observed case.}}
  \label{fig:titan_comm_parameters}
\end{figure}

>>>>

\subsection{Titan }
\label{subsec:titan}
All parallel scaling studies presented in this work were performed on
the Titan Cray XK7 machine at the Oak Ridge Leadership Computing
Facility located at Oak Ridge National
Laboratory\footnote{Specifications presented for Titan in the section
  were gathered from the Oak Ridge Leadership Computing Facility at
  \url{http://www.olcf.ornl.gov/computing-resources/titan-cray-xk7/}}. Titan
is a heterogeneous architecture machine combining both CPU and GPU
components to achieve a theoretical peak performance of more than 20
petaflops. The technical specifications for Titan are given in
Table~\ref{tab:titan_hardware} and spefically for only the CPU portion
of the machine.
\begin{table}[h!]
  \begin{center}
    \begin{tabular}{lc}\hline\hline
      Processor & 16 core AMD \\
      Nodes & 18,688 AMD Opterons \\
      Cores/Node & 16 \\
      Total Cores & 299,008 \\
      Memory/Node & 32 GB \\
      Memory/Core & 2 GB \\
      Interconnect & Gemini \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Titan Cray XK7 hardware specifications.}}
  \label{tab:titan_hardware}
\end{table}

For the scaling studies presented in this work, only the CPU
components specified in of Titan were utilized. This is due to the
fact that the implementations of the parallel MCSA algorithm utilized
code written only with the Message Passing Interface (MPI) and
therefore could not take advantage of the GPU accelerators. Even
considering just the CPU portion of titan, nearly 300,000 CPUs are
available to test the parallel performance of the algorithm at high
levels of concurrency.

For each parallel performance metric presented in the previous
section, a reference case is required to compute efficiencies for all
other cases. For smaller machines, often a serial computation can be
performed and used as the base case to get a true measure of speed-up
for a particular problem. For our scaling studies, we will always use
as a reference case a calculation which uses 16 cores, filling an
entire node of the machine. We fill an entire node so that we may
ignore the effects on scalibility measures that arise from using only
a subset of cores in a node. Typically this results in a large drop in
efficiency from a serial computation to a calculation that fills the
entire node. Once a node is filled, adding subsequent nodes has less
of an effect on the scaling, providing a more stable region of
analysis.

@@

In our description of Brunner and Brantley's algorithm as applied to
the Neumann-Ulam method, we noted two free parameters that are
user-defined at runtime. The first is the static data buffer size for
communicating histories between neighboring domains and the second is
the frequency by which a domain checks for incoming information from
neighboring domain, both of which may potentially affect latency and
load balancing. Changing the static history buffer size parameter
affects the scalability of the algorithm in two ways. If the buffer
size is increased, the buffer takes longer to fill and is therefore
sent less frequently, potentially causing scalability bottlenecks
where a neighboring domain is idle and waiting for more work to do but
is waiting for another domain to complete work. In addition, if the
buffer size is too large, the network may become saturated with these
very large data packets, slowing down performance. If the buffer size
is too small, too many history messages are sent, not providing enough
data for a neighboring process to work on to make up for the time
spent communicating. Considering the second parameter, when the check
frequency period is increased, more histories are processed locally
before looking to do work elsewhere. If this value becomes too large,
incoming data buffers may not be cleared frequently enough and cause a
bottle neck in the domain-to-domain communications. If this value is
too small, too much time is spent looking for work instead of actually
performing it.

To assess how these parameters affect the performance of the parallel
Monte Carlo algorithm on the Titan machine, a parameter study was
performed using the neutron diffusion problem on 64 cores (4
nodes). Each parameter was varied between 16 and 4096 and the time to
solve the problem recorded. The results of this study are presented in
Figure~\ref{fig:titan_comm_parameters}. Compared to the results of
Brunner and Brantley for the same type of study, we note qualitatively
similar results with the general observation that a larger history
buffer check period is better than a shorter one. The larger the check
period, the more on-process work is done before checking
messages. Second, there was not a significant variation observed in
the history buffer size parameter for a given check
frequency.

\begin{figure}[htpb!]
  \begin{center}
    \includegraphics[width=6in]{chapters/parallel_mc/titan_comm_parameters.pdf}
  \end{center}
  \caption{\textbf{Neumann Ulam parallel algorithm free parameter
      sensitivity.} \textit{The color scale represents the ratio of
      the runtime of a particular parameter variation with the runtime
      from the faster parameter variation. A value of 1.02 on the
      color scale corresponds to a runtime 2\% longer than the faster
      time. All runtimes were within 6.5\% of the fastest observed
      case.}}
  \label{fig:titan_comm_parameters}
\end{figure}

Compared to the reference work, there was also not a significant
change in the runtime over the values of parameters tested with all
runtimes reported within 6.5\% of the fastest case observed. For this
work, a history buffer size of 4096 and a buffer check period of 128
performed best for the intitial calculations. However, most values in
the range tested will give good performance. Brunner and Brantley
found that a value of 1024 was best for both parameters. For the
results observed here, these parameters were only 0.7\% slower than
the fastest case. It may be best practice to fix these parameters
internally at 1024 as both this work and the reference found these to
perform well several years apart on both different parallel machines
and for different transport sequences.
